Correcting for bias in high-throughput sequencing data

### 2011-08-23 response/revision_plan.tex
added frt-seq data
parameter sensitivity explored in the suplement
comments regarding parameter sensitivity

\documentclass{article}
\usepackage{dcj}

@@

\item Demonstrated the results are not particularity sensitive to the
complexity penalty, drawing a plot similar to the one is Supplementary Section
3, but varying the penalty rather than the number or reads.

\item Point out that the fact that the method is effective over diverse sets of
data argues that the results are not sensitive to any of the parameters used.

@@

\item (Maybe) Draw a curve showing how the performance of Hansen's brute-force
method changes as a function of the number of reads fed into it.

\item Make very clear that in there is no bias present in the data, a sparse or
empty model will be trained.

\item Include a figure from the genome browser showing an example uniformity of
coverage being increased over a gene or exon.

\item Use the MAQC set to demonstrate the following:
\begin{enumerate}
\item Increased agreement between bias corrected sequencing data and qPCR.
\item Increased agreement between replicates.
\end{enumerate}

>>>>

\documentclass{article}
\usepackage{dcj}
\usepackage{ulem}

@@

\item \sout{Demonstrated the results are not particularity sensitive to the
complexity penalty, drawing a plot similar to the one is Supplementary Section
3, but varying the penalty rather than the number or reads.}

\item \sout{Point out that the fact that the method is effective over diverse sets of
data argues that the results are not sensitive to any of the parameters used.}

@@

%\item (Maybe) Draw a curve showing how the performance of Hansen's brute-force
%method changes as a function of the number of reads fed into it.

% This is sort of implied by the FRT-seq data
\item Make very clear that in there is no bias present in the data, a sparse or
empty model will be trained.

\item \sout{Include a figure from the genome browser showing an example uniformity of
coverage being increased over a gene or exon.}

\item Use the MAQC set to demonstrate the following:
\begin{enumerate}
\item Increased agreement between bias corrected sequencing data and qPCR.
%\item Increased agreement between replicates. % Won't work here
\end{enumerate}

\item \sout{Examine the bias in new methods without a PCR-amplification step.}

\item \sout{Typos mentioned by reviewr 1.}

### 2011-08-23 seqbias-paper.tex
comments regarding parameter sensitivity

\begin{figure*}
\centerline{\includegraphics[width=0.9\textwidth]{freqs.eps}}
\caption{Nucleotide frequencies are plotted relative to the start (labeled
position 0) of each mapped read, respecting strand, and grouped by platform
(Illumina or ABI SOLiD).  The data sets plotted here are those used
for evaluation, listed in Table \ref{tab:datasets}.
The sequence is taken from the genomic context surrounding the read, so that -40
to -1, for example, fall outside the read sequence itself. The symmetrized
Kullback-Leibler divergence is used to summarize the difference in nucleotide
frequency compared to a fixed estimate of background nucleotide frequencies made
by sampling many positions near mapped reads.  Under the assumption that reads
are sampled uniformly from transcripts, each of the plots should be essentially
flat}
\label{fig:freqs}
\end{figure*}

@@

This suggests a natural scheme in which observations may be reweighted to
correct for bias.  First, define the \emph{sequence bias} $b_i$ at position $i$
as,
$$ b_i = \frac{ \Pr[ s_i ] } { \Pr[ s_i | m_i ] } $$

@@

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

@@

\begin{figure*}
\centerline{\includegraphics[width=1.00\textwidth]{kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frequencies expected under the assumption of uniform sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted read counts as well as after adjusting read counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

In this comparison, we measure how well the counts conform to a Poisson process.
The assumption of positional read counts following a Poisson distribution is
known to be a poor fit \citep{Srivastava2010}, but measuring the improvement in
the fit derived from correcting for bias remains a principled and easily
interpreted criterion.

We perform maximum-likelihood fitting of two models. In the null model, the
Poisson rate is fixed for each exon. That is, for position $j$ within
exon $i$, the rate is
$$ \lambda_{ij} = a_i $$
where $a_i$ is the parameter being fit. For comparison, we then fit a model in
which the rate is also proportional to the predicted bias coefficients:
$$ \lambda'_{ij} = a_i b_{ij} $$

@@

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive, and increases as uniformity is
increased, and negative when uniformity is decreased. Boxes are plotted to mark
the 25\%, 50\%, and 75\% quantiles, with whiskers extending to 1.5 times the
inter-quartile range (i.e., the span between the 25\% and 75\% quantiles), and
dots marking more extreme values.
}

>>>>

\begin{figure*}
\centerline{\includegraphics[width=0.85\textwidth]{freqs.eps}}
\caption{Nucleotide frequencies are plotted relative to the start (labeled
position 0) of each mapped read, respecting strand, and grouped by platform
(Illumina or ABI SOLiD).  The data sets plotted here are those used
for evaluation, listed in Table \ref{tab:datasets}.
The sequence is taken from the genomic context surrounding the read, so that -40
to -1, for example, fall outside the read sequence itself. The symmetrized
Kullback-Leibler divergence is used to summarize the difference in nucleotide
frequency compared to a fixed estimate of background nucleotide frequencies made
by sampling many positions near mapped reads.  Under the assumption that reads
are sampled uniformly from transcripts, each of the plots should be essentially
flat}
\label{fig:freqs}
\end{figure*}

@@

This suggests a natural scheme in which observations may be reweighted to
correct for bias.  First, define the \emph{sequence bias} $b_i$ at position $i$
as $ b_i = \Pr[ s_i ] /  \Pr[ s_i | m_i ]  $.

@@

\begin{figure}
\centerline{\includegraphics[width=0.30\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

Along with $p_{\text{max}}$ and $d_{\text{max}}$, the time required to train the
model is controlled by the number of reads used (i.e., $n$) and the number of
positions surrounding the read start to consider. Increasing any of these
parameters results in a solution that is no worse, and potentially better, at
the expense of increased training time.

The quality of the solution depends also on two other parameters: the standard
deviation at which background sequences are sampled, and the weight applied to
the penalty term of the BIC, yet it is not particularly sensitive to their values
(see, Supplementary Section 4). In all our evaluation, no special tuning of the
parameters was performed, suggesting it can be used effectively across data sets
without any intervention.

@@

\begin{figure*}
\centerline{\includegraphics[width=0.95\textwidth]{kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frequencies expected under the assumption of uniform sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted read counts as well as after adjusting read counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{uniformity.eps}}
\caption{Read counts (i.e., the number of reads starting at each position)
from the Mortazavi data set are plotted along the $3'$ UTR of Apoa2. The upper
plot shows unadjusted counts, the lower, counts adjusted by our method.}
\label{fig:unif}
\end{figure}

In this comparison, we measure how well the counts conform to a Poisson process.
The assumption of positional read counts following a Poisson distribution is
known to be a poor fit \citep{Srivastava2010}, but measuring the improvement in
the fit derived from correcting for bias remains a principled and easily
interpreted criterion. This increase in uniformity is illustrated in Figure
\ref{fig:unif}.

We perform maximum-likelihood fitting of two models. In the null model, the
Poisson rate is fixed for each exon. That is, for position $j$ within exon $i$,
the rate is $ \lambda_{ij} = a_i $ where $a_i$ is the parameter being fit. For
comparison, we then fit a model in which the rate is also proportional to the
predicted bias coefficients: $ \lambda'_{ij} = a_i b_{ij} $.

@@

\begin{figure}
\centerline{\includegraphics[width=0.30\textwidth]{pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive, and increases as uniformity is
increased, and negative when uniformity is decreased. Boxes are plotted to mark
the 25\%, 50\%, and 75\% quantiles, with whiskers extending to 1.5 times the
inter-quartile range (i.e., the span between the 25\% and 75\% quantiles), and
dots marking more extreme values.
}
